---
title: "Beginner_Intermediate_General_Analysis"
author: "Ashraf Alsinglawi"
date: "2024-09-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Required Libraries

```{r Load Libraries, include=FALSE}
library(tidyverse)
library(lubridate)
library(ggplot2)
library(dplyr)
library(reshape2)
```

## Load the Dataset
```{r Loading Data, include=FALSE}
data <- read_csv("healthcare_dataset.csv")
```
## Data structure and summary
```{r Data Summary, echo=FALSE}
glimpse(data)
```
As we see, there are 15 columns(variables) with 10000 rows. All seems to be with the correct data types (will check later)

What is needed here:
1. Change variables names.
2. Check dates data types
3. Think about new features to extract and add for deep analysis


```{r Data Summary, echo=FALSE}
summary(data)
```
# As we see, We have 10 character variables. Gender, Blood Type, Admission Type and Test Results could be categorical in addition to Age if we want to analyze age categories. 
# 3 numerical Age, Billing Amount and Room Number, the latter is not very informative, so we consider drop it or make it character. 
# 2 Date variables, from which we could extract other variables.
# Min Age is 18, Max is 85 and the mean is 51.45
# Min Billing amount is 1000, Max is 49996 and the Mean is 25517

It's good to see the distribution of age variable, histogram is perfect in this case:

```{r}
# Histogram for numerical variable
ggplot(data, aes(x = Age)) +
  geom_histogram(binwidth = 10, fill = "blue", color = "white") +
  labs(title = "Distribution of Age", x = "Age")
```
So, almost all 10-year category having very close number of patients with lesser extent for age between 15-25 category. 

Let's check if there are missing values in the tibble

```{r Missing Data, echo=FALSE}
# Check for missing values
colSums(is.na(data))
```
# Good news! there are no missing values in the dataset


Checking duplicates
```{r Duplicate data, echo=FALSE}
duplicates <- data[duplicated(data), ]
head(duplicates, 5)
```
No duplicated rows!

# Data Cleaning
# Rename some column names to get rid of space:

```{r Rename Columns, echo=FALSE}
data <- data %>% 
  rename(BloodType = `Blood Type`, MedicalCondition = `Medical Condition`, AdmDate = `Date of Admission`, Insurance = `Insurance Provider`, BillingAmount = `Billing Amount`, RoomNum = `Room Number`, AdmType = `Admission Type`, DischDate = `Discharge Date`, TestResults = `Test Results`)
```

# Convert categorical variables to factors
```{r Change Data Type, echo=FALSE}
data <- data %>%
  mutate(across(where(is.character), as.factor))
```

```{r Check Data Types, echo=FALSE}
glimpse(data)
```

## Change room number to catecorical
```{r Change RoomNum to factor, echo=FALSE}
data <- data %>%
  mutate(RoomNum = as.factor(RoomNum))
```

```{r Check data, echo=FALSE}
glimpse(data)
```


## Detect of outliers Billing Amount using IQR
```{r Calculate IQR, echo=FALSE}
Q1 <- quantile(data$BillingAmount, 0.25)
Q3 <- quantile(data$BillingAmount, 0.75)
IQR <- Q3 - Q1
```

```{r Check Outliers, echo=FALSE}
data_Outliers <- data %>%
  filter(BillingAmount < (Q1 - 1.5 * IQR) & BillingAmount > (Q3 + 1.5 * IQR))

```

No Outliers!

## Count occurrences of each medical condition 
```{r Count Medical Conditions, echo=FALSE}
data %>%
  count(MedicalCondition, sort = TRUE) %>%
  arrange(desc(n))
```


# Average billing amount per medical condition
```{r Average Billing By Condition, echo=FALSE}
AverageBillingByCondition <- data %>%
  group_by(MedicalCondition) %>%
  summarise(AverageBilling = mean(BillingAmount, na.rm = TRUE))

print(AverageBillingByCondition)
```
## Plotting average cost of each diagnosis
```{r Average Cost of each Diagnosis, echo=FALSE}
ggplot(AverageBillingByCondition, aes(x = reorder(MedicalCondition, -AverageBilling), y = AverageBilling)) +
  geom_bar(stat = "identity", fill = "salmon") +
  labs(x = "Medical Condition", y = "Average Billing Amount", title = "Average Billing Amount by Medical Condition") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


## Average billing amount per Hospital

```{r Average Billing by Hospital, echo=FALSE}
AverageBillingByHospital <- data %>%
  group_by(Hospital) %>%
  summarise(AverageBilling = mean(BillingAmount, na.rm = TRUE))

print(AverageBillingByHospital)
```

## Top 10 Average billing amount per Hospital
```{r Top 10 Hospitals of Average Billing , echo=FALSE}
Top10AverageBillingByHospital <- data %>%
  group_by(Hospital) %>%
  summarise(AverageBilling = mean(BillingAmount, na.rm = TRUE)) %>%
  arrange(desc(AverageBilling)) %>%
  slice_head(n = 10)

print(Top10AverageBillingByHospital)
```

## Plotting average cost of Top 10 hospital
```{r Top 10 Hospitals in Cost, echo=FALSE}
ggplot(Top10AverageBillingByHospital, aes(x = reorder(Hospital, -AverageBilling), y = AverageBilling)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(x = "Hospital", y = "Average Billing Amount", title = "Average Billing Amount of Top 10 Hospital") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Lowest 10 Average billing amount per Hospital
```{r Lowest10 Hospitals of Average Billing, echo=FALSE}
Low10AverageBillingByHospital <- data %>%
  group_by(Hospital) %>%
  summarise(AverageBilling = mean(BillingAmount, na.rm = TRUE)) %>%
  arrange(desc(AverageBilling)) %>%
  slice_tail(n = 10)

print(Low10AverageBillingByHospital)
```

## Plotting average cost of Lowest 10 hospital
```{r Lowest 10 Hospitals in Cost, echo=FALSE}
ggplot(Low10AverageBillingByHospital, aes(x = reorder(Hospital, -AverageBilling), y = AverageBilling)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(x = "Hospital", y = "Average Billing Amount", title = "Average Billing Amount of Lowest 10 Hospital") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```



## Feature engineering
### Calculate and Add Hospital Length of Stay

```{r Add LOS, echo=FALSE}
data <- data %>%
  mutate(LOS = as.numeric(DischDate - AdmDate))
```

```{r Add DayCost, echo=FALSE}
data <- data %>%
  mutate(DayCost = BillingAmount / LOS)
```



## Visualization

```{r Blood Type Distribution & Gender, echo=FALSE}
ggplot(data = data) +
geom_bar(mapping = aes(x = BloodType, fill = Gender))
```

```{r Admission Type Distribution & Gender, echo=FALSE}
ggplot(data = data) +
geom_bar(mapping = aes(x = AdmType, fill = Gender))
```

```{r Medical Condition Distribution & Gender, echo=FALSE}
ggplot(data = data) +
geom_bar(mapping = aes(x = MedicalCondition, fill = Gender))
```

```{r Medical Condition Distribution & Blood Type, echo=FALSE}
ggplot(data = data) +
geom_bar(mapping = aes(x = MedicalCondition, fill = BloodType))
```

```{r Admission Type Distribution & Medical Condition, echo=FALSE}
ggplot(data = data) +
geom_bar(mapping = aes(x = AdmType, fill = MedicalCondition))
```

```{r Admission Type  Distribution & Test Results, echo=FALSE}
ggplot(data = data) +
geom_bar(mapping = aes(x = AdmType, fill = TestResults))
```
## Correlation Analysis
```{r Correlation , echo=FALSE}
cor_data <- data %>%
  select(where(is.numeric)) %>%
  cor()

print(cor_data)
```

## Heat map in ggplot requires data to be in long format, so we use melt function
```{r Melting data frame, echo=FALSE}
data_melt <- melt(cor_data)
data_melt
```

## Correlation heatmap
```{r Correlation heatmap, echo=FALSE}
ggplot(data_melt, aes(Var1, Var2)) +                           # Create heatmap with ggplot2
  geom_tile(aes(fill = value))
```



## Including Plots


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
