---
title: "Beginner_Intermediate_General_Analysis"
author: "Ashraf Alsinglawi"
date: "2024-09-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Required Libraries

```{r Load Libraries, include=FALSE}
library(tidyverse)
library(lubridate)
library(ggplot2)
```

## Load the Dataset
```{r Loading Data, include=FALSE}
data <- read_csv("healthcare_dataset.csv")
```
## Data structure and summary
```{r Data Summary, echo=FALSE}
glimpse(data)
```
As we see, there are 15 columns(variables) with 10000 rows. All seems to be with the correct data types (will check later)

What is needed here:
1. Change variables names.
2. Check dates data types
3. Think about new features to extract and add for deep analysis


```{r Data Summary, echo=FALSE}
summary(data)
```
# As we see, We have 10 character variables. Gender, Blood Type, Admission Type and Test Results could be categorical in addition to Age if we want to analyze age categories. 
# 3 numerical Age, Billing Amount and Room Number, the latter is not very informative, so we consider drop it or make it character. 
# 2 Date variables, from which we could extract other variables.
# Min Age is 18, Max is 85 and the mean is 51.45
# Min Billing amount is 1000, Max is 49996 and the Mean is 25517

It's good to see the distribution of age variable, histogram is perfect in this case:

```{r}
# Histogram for numerical variable
ggplot(data, aes(x = Age)) +
  geom_histogram(binwidth = 10, fill = "blue", color = "white") +
  labs(title = "Distribution of Age", x = "Age")
```
So, almost all 10-year category having very close number of patients with lesser extent for age between 15-25 category. 

Let's check if there are missing data in the tibble

```{r Missing Data, echo=FALSE}
# Check for missing values
colSums(is.na(data))
```
# Good news! there are no missing values in the dataset


## Including Plots


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
